Hypercoercions and a Framework for \\ Equivalence of Cast Calculi

### 2019-12-29 paper.tex
review according to most reviews

\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\mathtt{Unit}}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{unit}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{(#1 \; #2)}
\newcommand{\eOOcons}[4]{\mathtt{cons}^{\POOprod{#1}{#2}} \; #3 \; #4}
\newcommand{\eOOcar}[1]{\mathtt{fst} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{snd} \; #1}
\newcommand{\eOOinl}[3]{\mathtt{inl}^{\POOsum{#1}{#2}} \; #3}
\newcommand{\eOOinr}[3]{\mathtt{inr}^{\POOsum{#1}{#2}} \; #3}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \Rightarrow^{#2} #3}
\newcommand{\oOOinj}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{unit}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
%\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
% Jeremy prefers the following
\newcommand{\vOOfun}[3]{\langle \lambda  #2. #3, #1 \rangle}
\newcommand{\vOOtt}[0]{\mathtt{unit}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: D and
UD~\citep{siek2009exploring}. We are interested in the D blame
tracking strategy because it comes with a more straightforward notion of
safe cast compared to UD~\citep{siek2009exploring}, which is why D was
chosen in the Grift compiler~\citep{kuhlenschmidt2018efficient}. We are also 
interested in UD because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The UD hypercoercions were inspired by the
supercoercions of \citet{Garcia:2013:CTB:2500365.2500603} (hence the name) and
the D hypercoercions were inspired by the normal forms of
\citet{siek2012interpretations}.
%
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper, we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that the eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
machine and Fig.~\ref{fig:state} gives a grammar for machine states
$s$, including a definition of values $v$ and value typing. 
%Fig.~\ref{fig:state} gives a grammar for machine states
%$s$, including a definition of values and value typing and
%Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
%machine.
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions, (1) those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation, and (2) those that return a value to the
current continuation and possibly perform a computation.
Corresponding to (1) and (2), the machine state is either in an evaluating
$\sOOinspect{e}{\mathcal{E}}{\kappa}$ or continuing $\sOOreturn{v}{\kappa}$ configuration, 
respectively. Additionally, there is the $\sOOhalt{o}$ configuration, where the 
machine halts with an observable ($o$). Observables include all value 
constructors and blame. The function converting values to 
observables ($observe(v) = o$) is defined in the obvious way.

@@

\begin{figure}
  Machine state and other runtime data structures
  \[
  \begin{array}{lrcl}
  \stxrule{\mathcal{E}}{Environments}{
  	\text{a partial function}\;\{\langle x , v \rangle , \dots \}
%  	\text{partial functions that map  variables to values}
  }
  \stxrule{v}{Values}{
    \vOOtt{} \mid
    \vOOfun{\mathcal{E}}{x}{e} \mid
    \vOOcons{v}{v} \mid
    \vOOinl{v} \mid
    \vOOinr{v} \mid   
    \vOOcast{v}{c}
  }
  \stxrule{c}{Casts}{
    \cOOcast{T}{l}{T}
  }
  \stxrule{I}{Injectable types (\lazyD)}{
    P
  }
  \stxrule{I}{Injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\star}{\star} \mid
    \star \times \star \mid
    \star + \star
  }
  \stxrule{o}{Observables}{
    \oOOinj \mid
    \oOOsole \mid
    \oOOfun \mid
    \oOOcons \mid
    \oOOinl \mid
    \oOOinr \mid
    \oOOblame{l}
  }
  \stxrule{r}{Cast results}{
    \rOOsucc{v} \mid
    \rOOfail{l}
  }
  \stxrule{s}{States}{
    \sOOinspect{e}{\mathcal{E}}{\kappa} \mid{}
    \sOOreturn{v}{\kappa} \mid{}
    \sOOhalt{o}
  }
  \stxrule{\kappa}{Continuations}{
    \kOOmt \mid
    \kOOconsI{T}{T}{e}{\mathcal{E}}{\kappa} \mid
    \kOOconsII{T}{T}{v}{\kappa} \mid
    \kOOinl{T}{T}{\kappa}
  }
  \stxrulecont{
    \kOOinr{T}{T}{\kappa} \mid
    \kOOappI{e}{\mathcal{E}}{\kappa} \mid
    \kOOappII{v}{\kappa} \mid
    \kOOcar{\kappa}
  }
  \stxrulecont{ 
    \kOOcdr{\kappa}\mid
    \kOOcaseI{e}{e}{\mathcal{E}}{\kappa}\mid
    \kOOcaseII{v}{e}{\mathcal{E}}{\kappa}
  }
  \stxrulecont{
    \kOOcaseIII{v}{v}{\kappa} \mid
    \kOOcast{c}{\kappa}
  }
  \end{array}
  \]

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observable that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercoercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the no-op,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the no-op.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.

>>>>

\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\mathtt{Unit}}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{unit}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{(#1 \; #2)}
\newcommand{\eOOcons}[4]{\mathtt{cons}^{\POOprod{#1}{#2}} \; #3 \; #4}
\newcommand{\eOOcar}[1]{\mathtt{fst} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{snd} \; #1}
\newcommand{\eOOinl}[3]{\mathtt{inl}^{\POOsum{#1}{#2}} \; #3}
\newcommand{\eOOinr}[3]{\mathtt{inr}^{\POOsum{#1}{#2}} \; #3}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \Rightarrow^{#2} #3}
\newcommand{\oOOdyn}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{unit}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
%\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
% Jeremy prefers the following
\newcommand{\vOOfun}[3]{\langle \lambda  #2. #3, #1 \rangle}
\newcommand{\vOOtt}[0]{\mathtt{unit}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: D and
UD~\citep{siek2009exploring}. With the D strategy, only \textbf{d}own casts 
(casts from $\TOOdyn$ to some other types) are subject to blame. With the UD 
strategy, however, some \textbf{u}p casts are also blamable.
We are interested in the D blame
tracking strategy because it comes with a more straightforward notion of
safe cast compared to UD~\citep{siek2009exploring}, which is why D was
chosen in the Grift compiler~\citep{kuhlenschmidt2018efficient}. We are also 
interested in UD because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The UD hypercoercions were inspired by the
supercoercions of \citet{Garcia:2013:CTB:2500365.2500603} (hence the name) and
the D hypercoercions were inspired by the normal forms of
\citet{siek2012interpretations}.
%
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper, we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that the eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
machine and Fig.~\ref{fig:state} gives a grammar for machine states
$s$, including a definition of values $v$ and value typing. 
%Fig.~\ref{fig:state} gives a grammar for machine states
%$s$, including a definition of values and value typing and
%Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
%machine.
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions, (1) those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation, and (2) those that return a value to the
current continuation and possibly perform a computation.
Corresponding to (1) and (2), the machine state is either in an evaluating
$\sOOinspect{e}{\mathcal{E}}{\kappa}$ or continuing $\sOOreturn{v}{\kappa}$ configuration, 
respectively. Additionally, there is the $\sOOhalt{o}$ configuration, where the 
machine halts with an observable ($o$). Observables include all value 
constructors, blame, and $\oOOdyn$, as in \citet{siek2012interpretations}. 
The function converting values to observables ($observe(v) = o$) is defined in 
the obvious way.

@@

\begin{figure}
  Machine state and other runtime data structures
  \[
  \begin{array}{lrcl}
  \stxrule{\mathcal{E}}{Environments}{
  	\text{a partial function}\;\{\langle x , v \rangle , \dots \}
%  	\text{partial functions that map  variables to values}
  }
  \stxrule{v}{Values}{
    \vOOtt{} \mid
    \vOOfun{\mathcal{E}}{x}{e} \mid
    \vOOcons{v}{v} \mid
    \vOOinl{v} \mid
    \vOOinr{v} \mid   
    \vOOcast{v}{c}
  }
  \stxrule{c}{Casts}{
    \cOOcast{T}{l}{T}
  }
  \stxrule{I}{Injectable types (\lazyD)}{
    P
  }
  \stxrule{I}{Injectable types (\lazyUD)}{
    \POOunit \mid
    \POOfun{\star}{\star} \mid
    \star \times \star \mid
    \star + \star
  }
  \stxrule{o}{Observables}{
    \oOOdyn \mid
    \oOOsole \mid
    \oOOfun \mid
    \oOOcons \mid
    \oOOinl \mid
    \oOOinr \mid
    \oOOblame{l}
  }
  \stxrule{r}{Cast results}{
    \rOOsucc{v} \mid
    \rOOfail{l}
  }
  \stxrule{s}{States}{
    \sOOinspect{e}{\mathcal{E}}{\kappa} \mid{}
    \sOOreturn{v}{\kappa} \mid{}
    \sOOhalt{o}
  }
  \stxrule{\kappa}{Continuations}{
    \kOOmt \mid
    \kOOconsI{T}{T}{e}{\mathcal{E}}{\kappa} \mid
    \kOOconsII{T}{T}{v}{\kappa} \mid
    \kOOinl{T}{T}{\kappa}
  }
  \stxrulecont{
    \kOOinr{T}{T}{\kappa} \mid
    \kOOappI{e}{\mathcal{E}}{\kappa} \mid
    \kOOappII{v}{\kappa} \mid
    \kOOcar{\kappa}
  }
  \stxrulecont{ 
    \kOOcdr{\kappa}\mid
    \kOOcaseI{e}{e}{\mathcal{E}}{\kappa}\mid
    \kOOcaseII{v}{e}{\mathcal{E}}{\kappa}
  }
  \stxrulecont{
    \kOOcaseIII{v}{v}{\kappa} \mid
    \kOOcast{c}{\kappa}
  }
  \end{array}
  \]

@@

This section presents our first contribution, the definition of
hypercoercions.  The design of hypercoercions is motivated by the
observation that a normal coercion has at most three
parts. Hypercoercions make this structure explicit: a hypercoercion $c$ is
either $\hyperCoercionI$, the identity cast for $\TOOdyn$, or it contains
three parts: a head, middle, and tail, as defined in
\figref{fig:hypercoercion}.
\begin{itemize}
\item The head $h$ is either a projection or the no-op,
\item the middle $m$ involves coercions that preserve a type
  constructor, i.e., coercions between function, pair, sum types, and
  unit types, and
\item the tail $t$ is either an injection, a failure, or the no-op.
\end{itemize}
There is a close correspondence between middle coercions and type
constructors.  We generalize shallow-consistency to middles $m \smile
m$ in the obvious way.
